
---
title: "FitBit Data Analysis"
author: "Evan Cutler"
date: "Feb, 2017"
---
###About
The purpose of the project was to answer a series of questions using data collected from a [FitBit](http://en.wikipedia.org/wiki/Fitbit).


##Synopsis
practice elements for this project:

* consuming data
* filling in missing values
* answer research questions from analyzing finalized datasets.

## Data
Original location of data:

* Dataset: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) [52K]

The variables included in this dataset are:

* **steps**: Number of steps taking in a 5-minute interval (missing
    values are coded as `NA`)

* **date**: The date on which the measurement was taken in YYYY-MM-DD
    format

* **interval**: Identifier for the 5-minute interval in which
    measurement was taken

The dataset comes from a comma-separated-value (CSV) file and there are 17,568 observations in this dataset.

## Loading and preprocessing the data

Download, unzip and load data into data frame `data`. 
```{r}
if(!file.exists("getdata-projectfiles-UCI HAR Dataset.zip")) {
        temp <- tempfile()
        download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",temp)
        unzip(temp)
        unlink(temp)
}

data <- read.csv("activity.csv")
```


## What is mean total number of steps taken per day?
Sum steps by day, create Histogram, and calculate mean and median.
```{r} 
steps_by_day <- aggregate(steps ~ date, data, sum)
hist(steps_by_day$steps, main = paste("Total Steps Each Day"), col="blue", xlab="Number of Steps")
rmean <- mean(steps_by_day$steps)
rmedian <- median(steps_by_day$steps)
```

The `mean` is `r rmean` and the `median` is `r rmedian`.

## What is the average daily activity pattern?

* find average steps for each interval for all days. 
* Plot Average Number Steps per Day by Interval. 
* Find interval with most average steps. 
```{r}
aggr_steps <- aggregate(steps ~ interval, data, mean)

plot(aggr_steps$interval,aggr_steps$steps, type="l", xlab="Interval", ylab="Number of Steps",main="Average Number of Steps per Day by Interval")

max_interval <- aggr_steps[which.max(aggr_steps$steps),1]
```

The 5-minute interval, on average across all the days in the data set, containing the maximum number of steps is `r max_interval`.

## Impute missing values. Compare data with missing values against data where missing data is repaired.
Missing data needed to be repaired (Inputed) Only a simple approach was required for this assignment. 
Missing values were repaired and the dataset was finalized by inserting the average for each interval. Thus, if interval 10 was missing on 10-02-2012, the average for that interval for all days (0.1320755), replaced the NA. 
```{r}
incomplete <- sum(!complete.cases(data))
finalized_data <- transform(data, steps = ifelse(is.na(data$steps), aggr_steps$steps[match(data$interval, aggr_steps$interval)], data$steps))
```


Recount total steps by day and create Histogram. 
```{r}
steps_by_day_i <- aggregate(steps ~ date, finalized_data, sum)
hist(steps_by_day_i$steps, main = paste("Total Steps Each Day"), col="blue", xlab="Number of Steps")

#Create Histogram to show difference. 
hist(steps_by_day$steps, main = paste("Total Steps Each Day"), col="red", xlab="Number of Steps", add=T)
legend("topright", c("Imputed", "Non-imputed"), col=c("blue", "red"), lwd=10)
```

Calculate new mean and median for imputed data. 
```{r}
rmean.i <- mean(steps_by_day_i$steps)
rmedian.i <- median(steps_by_day_i$steps)
```

Calculate difference between imputed and non-imputed data.
```{r}
mean_diff <- rmean.i - rmean
med_diff <- rmedian.i - rmedian
```

Calculate total difference.
```{r}
total_diff <- sum(steps_by_day_i$steps) - sum(steps_by_day$steps)
```
* The imputed data mean is `r rmean.i`
* The imputed data median is `r rmedian.i`
* The difference between the non-imputed mean and imputed mean is `r mean_diff`
* The difference between the non-imputed mean and imputed mean is `r med_diff`
* The difference between total number of steps between imputed and non-imputed data is `r total_diff`. Thus, there were `r total_diff` more steps in the imputed data.


## Are there differences in activity patterns between weekdays and weekends?
Created a plot to compare and contrast number of steps between the week and weekend. There is a higher peak earlier on weekdays, and more overall activity on weekends.  
``` {r}
library(lattice)
weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
dow = as.factor(ifelse(is.element(weekdays(as.Date(finalized_data$date)),weekdays), "Weekday", "Weekend"))

aggr_steps_i <- aggregate(steps ~ interval + dow, finalized_data, mean)


xyplot(aggr_steps_i$steps ~ aggr_steps_i$interval|aggr_steps_i$dow, main="Average Steps per Day by Interval",xlab="Interval", ylab="Steps",layout=c(1,2), type="l")

```